---
title: "Hand measurements"
author: "Philipp Wacker"
date: "2024-07-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load libraries

```{r}
library(ggplot2)
```

## Hand measurements

```{r}
span <- c(19, 18, 21, 20.7, 18.5, 21.5, 22)
cubit <- c(45, 45.5, 46.2, 46, 39, 43, 49)
df_hand <- data.frame(span = span, cubit = cubit)
write.csv(df_hand, file="measurements.csv")
```

Compute the mean handspan by

$mean(span) = \frac{1}{N}\sum_{i=1}^N s_i$

```{r}
mean_span <- sum(span)/length(span)
```

Next we compute the variance

$var(span) = \frac{1}{N-1}\sum_{i=1}^N (s_i - mean(span))^2$

```{r}
var_span <- sum((span - mean_span)^2) /( length(span) - 1)
sd_span <- sqrt(var_span)
```

Plot `span` vs `cubit`:

```{r}
ggplot(data=df_hand, aes(x=cubit, y=span)) +
  geom_point()
```

Investigate correlation between these variables

```{r}
cov_span_cubit <- NULL
corr_span_cubit <- NULL
```

## Building confidence intervals for the estimator $\hat \mu$ of handspan

We are going to estimate the mean handspan with the unbiased mean estimator

$$ \hat \mu = \frac{1}{n} \sum_{i=1}^n x_i. $$

A confidence interval for $\mu$ at confidence level $1-\alpha=$ 95% is computed via

$$ \left[\hat \mu - \underbrace{t_{1-\alpha/2}(n-1) \cdot \frac{\sqrt{ \widehat{\sigma^2}}}{\sqrt n}}_{=:~ \tau},~ \hat \mu + t_{1-\alpha/2}(n-1) \cdot \frac{\sqrt{ \widehat{\sigma^2}}}{\sqrt n}  \right]  $$
$$[ \hat \mu - t(1-\alpha/2, n-1)*\sqrt{\hat \sigma^2/n}, \hat \mu + t(1-\alpha/2, n-1)*\sqrt{\hat \sigma^2/n}]   $$

```{r}
alpha = 0.02
n <- nrow(df_hand)
hatmu <- mean_span
tau <- qt(1-alpha/2, n-1)*sqrt(var_span)/sqrt(n)
print(c(hatmu - tau, hatmu + tau))
```
The `r 100*(1-alpha)`% confidence interval for $\mu$ is [`r signif(c(hatmu - tau, hatmu + tau), 4)`]

```{r fig.height=2, fig.width=6}
ggplot(data=df_hand) +
  geom_point(aes(x=span, y=replicate(length(span), 0))) + 
  geom_point(aes(x=hatmu, y=0), color="red", size=3) +
  geom_errorbarh(aes(xmin = hatmu-tau, xmax = hatmu+tau, y=0))
```

